---
title: "Session_3_Obesity"
output: html_document
date: "2025-03-05"
---


# Install packages

```{r}

install.packages("classInt")
install.packages("curl")
install.packages("dbscan")
install.packages("downloader")
install.packages("dplyr")
install.packages("fasterize")
install.packages("foreign")
install.packages("fpc")
install.packages("fs")
install.packages("geojsonio")
install.packages("ggplot2")
install.packages("h3jsr")
install.packages("h3r")
install.packages("here")
install.packages("janitor")
install.packages("osmdata")
install.packages("osmextract")
install.packages("plotly")
install.packages("r5r")
install.packages("raster")
install.packages("readxl")
install.packages("RColorBrewer")
install.packages("sf")
install.packages("sp")
install.packages("spData")
install.packages("spatstat")
install.packages("spatstat.geom")
install.packages("stars")
install.packages("terra")
install.packages("tidyterra")
install.packages("tidyverse")
install.packages("tmap")
install.packages("tmaptools")

```

# Load packages

```{r}

library(classInt)
library(curl)
library(dbscan)
library(downloader)
library(dplyr)
library(fasterize)
library(foreign)
library(fpc)
library(fs)
library(geojsonio)
library(ggplot2)
library(h3jsr)
library(h3r)
library(here)
library(janitor)
library(osmdata)
library(osmextract)
library(plotly)
library(r5r)
library(raster)
library(readxl)
library(RColorBrewer)
library(sf)
library(sp)
library(spData)
library(spatstat)
library(spatstat.geom)
library(stars)
library(terra)
library(tidyterra)
library(tidyverse)
library(tmap)
library(tmaptools)

```

# Import necessary data
## Read in shapefile of London

```{r}

# Define the URL of the ZIP file
url <- "https://data.london.gov.uk/download/statistical-gis-boundary-files-london/9ba8c833-6370-4b11-abdc-314aa020d5e0/statistical-gis-boundaries-london.zip"

# Function to download, extract, and load a shapefile
load_shapefile <- function(zip_url, shapefile_relative_path) {
  temp <- tempfile(fileext = ".zip")
  curl_download(zip_url, temp)
  
  temp_dir <- tempfile()
  dir.create(temp_dir)
  unzip(temp, exdir = temp_dir)
  
  shapefile_path <- file.path(temp_dir, shapefile_relative_path)
  
  if (!file.exists(shapefile_path)) {
    stop(paste("Shapefile not found:", shapefile_relative_path))
  }
  
  return(st_read(shapefile_path))
}

# Load London Boroughs shapefile
london_boroughs <- load_shapefile(url, "statistical-gis-boundaries-london/ESRI/London_Borough_Excluding_MHW.shp") %>% 
st_transform(london_boroughs, crs = 4326)
qtm(london_boroughs)

# Load LSOA shapefile
london_LSOA <- load_shapefile(url, "statistical-gis-boundaries-london/ESRI/LSOA_2011_London_gen_MHW.shp") %>% 
st_transform(london_boroughs, crs = 4326)
qtm(london_LSOA)

```

## Download and read in the Index of Multiple Deprivation (IMD) data from the CDRC

```{r}

#Load Index of Multiple Deprivation data. 
#Downloaded data from the CDRC Index of Multiple Deprivation (IMD) website: https://data.cdrc.ac.uk/dataset/index-multiple-deprivation-imd/resource/data-english-imd-2019
#English IMD 2019. 2011 LSOA geography. Shapefile format.

IMD_London <- st_read(here::here("English IMD 2019", "IMD_2019.shp"))%>%
  clean_names()%>%
  dplyr::filter(str_detect(la_dcd, "^E09"))

#Plot to see IMD deciles and check if all is correct
tmap_options(check.and.fix = TRUE)

tmap_mode("plot")

#Plot map
map <- tm_shape(IMD_London) +
  tm_polygons(col = "IMD_Decile", 
              style = "pretty", 
              n = 10, 
              palette = "RdYlBu", 
              title = "IMD Decile", 
              border.alpha = 0) +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_shape(lnd_ff_pts) +
  tm_dots(col = "black",
          size = 0.05,   
          border.lwd = 0, 
          title = "Fast-food outlets",  # Label for the dots
          legend.show = TRUE)  

print(map)

```

## Load Fast-food points from Open Street Map (OSM)

```{r}

# Define the geographic area (example: Greater London)
Bounding_box <- getbb("Greater London")  # Bounding box for London

# Query OSM for fast food locations
fast_food <- opq(Bounding_box) %>%
  add_osm_feature(key = "amenity", value = "fast_food") %>%
  osmdata_sf()

# Extract point locations (fast food outlets)
fast_food_points <- fast_food$osm_points %>% 
  st_transform(4326) %>% 
  mutate(opps = 1)

# View first few rows
head(fast_food_points)

#Filter out the points outside of our study area
fast_food_points <- fast_food_points[london_boroughs,]

# Plot fast food outlets
plot(st_geometry(fast_food_points), col = "red", pch = 20, main = "Fast Food Outlets in London")

```

## Load school points from Open Street Map (OSM)

```{r}

# Define the geographic area (example: Greater London)
Bounding_box <- getbb("Greater London")  # Bounding box for London

# Query OSM for school locations
schools <- opq(Bounding_box) %>%
  add_osm_feature(key = "amenity", value = "college") %>%
  osmdata_sf()

# Extract point locations (schools)
schools_points <- schools$osm_points %>% 
  st_transform(4326) %>% 
  mutate(opps = 1)

# View first few rows
head(schools_points)

#Filter out the schools outside of our study area
schools_points <- schools_points[london_boroughs,]

# Plot schools
plot(st_geometry(schools_points), col = "blue", pch = 20, main = "Schools in London")

```

## Creating Point Pattern Process (PPP) Objects

```{r}

# Project schools and fast food outlets to British National Grid CRS
lnd_sch_bng <- st_transform(schools_points, 27700)
lnd_ff_bng <- st_transform(fast_food_points, 27700)
lnd_bng <- st_transform(london_boroughs, 27700)

# Summarize the transformed school points
summary(lnd_sch_bng)

# Convert schools and fast food locations to PPP objects for spatial analysis
lnd_sch_ppp <- as.ppp(lnd_sch_bng)
lnd_ff_ppp <- as.ppp(lnd_ff_bng)

# Remove marks (attributes) from the PPP objects
marks(lnd_sch_ppp) <- NULL
marks(lnd_ff_ppp) <- NULL

# Convert London borough boundaries to observation window format
lnd_owin <- as.owin(lnd_bng)

# Assign the observation window to the fast food PPP object
Window(lnd_ff_ppp) <- lnd_owin

# Plot the fast food point pattern objects
plot(lnd_ff_ppp)

```

# Part 1 - IMD (Index of Multiple Deprivation) Analysis

## The aim of this analysis is to investigate whether areas with higher levels of deprivation have a greater density of fast-food outlets. This will help us determine if there are inequalities in fast-food outlet distribution based on area level deprivation - and if so, in which areas public health interventions should be prioritised.  

## Creating a Raster from IMD Data
```{r}

# Convert the IMD shapefile into a raster format for analysis
IMD_London_raster <- st_rasterize(IMD_London %>% dplyr::select(IMD_Decile, geometry))

# Plot the generated raster
plot(IMD_London_raster)

# Convert the raster to an IM object for statistical analysis
IMD_London_im <- as.im(IMD_London_raster)

# Plot the IM object
plot(IMD_London_im)

```

## Quadrat Analysis on Fast Food Locations
```{r}

# Set up the graphical parameters for a better layout
par(mar = c(4, 4, 2, 2))  # Adjust margins for visualization

dev.new()  # Open a new plotting window to visualize the quadrat analysis

# Perform quadrat count analysis to assess spatial distribution
quad_ff_IMD <- quadratcount(lnd_ff_ppp, tess = IMD_London_im)

# Display quadrat count results
quad_ff_IMD

# Plot the quadrat count analysis
plot(quad_ff_IMD)

```

## Density Estimation with Rho Hat Plot
```{r}

# Generate and plot the rho-hat function to assess intensity variation by area level deprivation
plot(rhohat(lnd_ff_ppp, IMD_London_im))

```

#Part 2 - Extending Spatial Point Pattern Analysis in the presence of a covariate (buffers around schools)

## The aim of this analysis is to determine which radius from schools captures the highest density of fast-food outlets. This will help us answer the question "How big should school fast-food exclusion zones be to cover areas of greatest exposure for school children?"

```{r}

# Create buffers around the schools
buffer_400m <- st_buffer(schools_points, dist = 400)
buffer_800m <- st_buffer(schools_points, dist = 800)
buffer_1200m <- st_buffer(schools_points, dist = 1200)

# Optionally, you can combine these buffers into a single object to visualize them
buffers_combined <- rbind(buffer_400m, buffer_800m, buffer_1200m)

# Plot the buffers to visualize the areas around the schools
plot(st_geometry(buffers_combined), col = c("red", "green", "blue"), main = "Buffers Around Schools", pch = 20)

# Now, let's rasterize each of the buffers using an appropriate rasterization process
London_boroughs_raster <- st_rasterize(london_boroughs %>% dplyr::select(NAME, geometry))

# Plot the rasterized IMD data
plot(London_boroughs_raster)

# Rasterize each buffer (this can be done for each individual buffer as well)
buffer_400m_raster <- st_rasterize(buffer_400m)
buffer_800m_raster <- st_rasterize(buffer_800m)
buffer_1200m_raster <- st_rasterize(buffer_1200m)



# Adam - this is as far as I've gotten with the buffers. I'm not sure how to rasterize them. Do I need to rasterize the map and then the buffers seperately and then join them? Would be grateful if you could please help with this section of the code. 



# Plot the rasterized buffers to see the surface area coverage
plot(buffer_400m_raster, main = "400m Buffer Raster")
plot(buffer_800m_raster, main = "800m Buffer Raster")
plot(buffer_1200m_raster, main = "1200m Buffer Raster")

# Create an IM object from the raster (you can do this for any of the buffers)
buffer_400m_im <- as.im(buffer_400m_raster)
buffer_800m_im <- as.im(buffer_800m_raster)
buffer_1200m_im <- as.im(buffer_1200m_raster)

# Plot the IM objects
plot(buffer_400m_im, main = "400m Buffer IM")
plot(buffer_800m_im, main = "800m Buffer IM")
plot(buffer_1200m_im, main = "1200m Buffer IM")

```

# Part 3 - Extending Spatial Point Pattern Analysis in the presence of a covariate (walking time bands) using R5 accessibility analysis

## The aim of this analysis is to determine the walking distance from schools that captures the highest density of fast-food outlets. This will help us answer the question "Within what walking distance from schools are children the most exposed to fast-food outlets?" We are going to do this for a single London borough for the purpose of this exercise (the borough of Croydon)

## Setup R5 Environment
```{r}

# Initialize Java Virtual Machine
rJava::.jinit()
rJava::.jcall("java.lang.System", "S", "getProperty", "java.version")
options(java.parameters = "-Xmx50G")  # Set memory allocation

# Load OpenStreetMap data for London
oe_match("London, England")
roads_london = oe_get("London, England", stringsAsFactors = FALSE, quiet = TRUE)
names(roads_london)
summary(roads_london)

# Filter major road types
ht = c("motorway", "trunk", "primary", "secondary", "tertiary", "residential", "unclassified")
osm_London_maj_roads = roads_london[roads_london$highway %in% ht, ]
plot(osm_London_maj_roads["highway"], key.pos = 1)
plot(sf::st_geometry(roads_london))

# Setup R5 transport network
here()
r5r_core <- setup_r5(here())

```

## Reproject Data to a Consistent CRS (EPSG 4326)
```{r}

# Convert spatial data to EPSG 4326
school_points_EPSG <- schools_points %>% 
  st_transform(4326) %>% 
  mutate(opps = 1) %>% 
  mutate(id = row_number())

fastfood_points_EPSG <- fast_food_points %>% 
  st_transform(4326) %>% 
  mutate(opps = 1)

london_boroughs_EPSG <- london_boroughs %>% 
  st_transform(4326)

```

## Filter Data for the Borough of Croydon Only
```{r}

# Extract Croydon boundary
croydon <- london_boroughs_EPSG %>% 
  filter(GSS_CODE == "E09000008")

# Extract Croydon-specific school and fast food locations
croydon_sch_pts <- school_points_EPSG[croydon,]
croydon_ff_pts <- fastfood_points_EPSG[croydon,]

```

## Generate Hexagonal Grid for Croydon
```{r}

croydon_h3 <- polygon_to_cells(croydon, res = 11, simple = FALSE)
croydon_h3 <- cell_to_polygon(unlist(croydon_h3$h3_addresses), simple = FALSE)
croydon_h3_centroids <- st_centroid(croydon_h3)

# Summarize and visualize the hexagonal grid
summary(croydon_h3)

ggplot() +
  geom_sf(data = croydon, fill = NA) +
  geom_sf(data = croydon_h3, fill = NA, colour = 'red') +
  ggtitle('Resolution 10 hexagons', subtitle = 'Croydon') +
  theme_minimal() +
  coord_sf()

tm_shape(croydon) +
  tm_polygons(col = "red", alpha = 0) +
tm_shape(croydon_h3) +
  tm_polygons(col = NA, alpha = 0) +
tm_shape(croydon_ff_pts) +
  tm_dots(col = "blue", border.lwd = 0) + 
tm_shape(croydon_sch_pts) +
  tm_dots(col = "black", border.lwd = 0) 

```

## Travel Time Matrix Calculation
```{r}

# Prepare data for travel time calculations
croydon_sch = croydon_sch_pts %>% st_drop_geometry()
croydon_h3_centroids[c("x", "y")] = st_coordinates(croydon_h3_centroids)

# Assign IDs
croydon_h3_centroids <- mutate(croydon_h3_centroids, id = row_number())
croydon_h3 <- mutate(croydon_h3, id = row_number())

# Define travel parameters
mode = c("WALK")
max_walk_time = 30  # Maximum walking time in minutes
max_trip_duration = 120  # Maximum trip duration in minutes
departure_datetime = as.POSIXct("01-12-2022 8:30:00", format = "%d-%m-%Y %H:%M:%S")

# Compute travel time matrix
ttm_croydon_h3 = travel_time_matrix(
  r5r_core = r5r_core,
  origins = croydon_sch_pts,
  destinations = croydon_h3_centroids,
  mode = mode,
  departure_datetime = departure_datetime,
  max_walk_time = max_walk_time,
  max_trip_duration = max_trip_duration
)

head(ttm_croydon_h3)
nrow(ttm_croydon_h3)
summary(ttm_croydon_h3)

```

## Identify Closest Schools by Travel Time
```{r}

# Find minimum travel time per destination
closest = aggregate(ttm_croydon_h3$travel_time_p50, by = list(ttm_croydon_h3$to_id), FUN = min, na.rm = TRUE)
summary(closest)

# Rename columns
closest <- rename(closest, id = Group.1, time = x)
closest["id"] = as.integer(closest$id)
head(closest)

# Join spatial data with travel time results
geo = inner_join(croydon_h3_centroids, closest, by = "id")
geo_hex = inner_join(croydon_h3, closest, by = "id")
head(geo)

```

## Visualizing Travel Time Data
```{r}

tm_shape(geo) +  
  tm_symbols(col = "time", size = 0.05, border.lwd = 0, style = "pretty", n = 10, palette = "RdYlBu", alpha = 0.3) +  
tm_shape(croydon_sch_pts) +
  tm_dots(col = "black", border.lwd = 0) +  
tm_shape(croydon) +
  tm_polygons(alpha = 0)

```

## Convert Data to British National Grid (EPSG 27700) and Analyze Spatial Distribution
```{r}

# Reproject spatial data
geo_hex_bng <- st_transform(geo_hex, 27700)

# Convert raster data
raster_geo <- st_rasterize(geo_hex_bng %>% dplyr::select(time, geometry)) 
st_coordinates(raster_geo)
plot(raster_geo, axes = T)
geo_hex_im <- as.im(raster_geo)
plot(geo_hex_im)

```

## Quadrat Analysis of Fast Food Locations
```{r}

# Convert to spatial point pattern objects
croy_sch_bng <- st_transform(croydon_sch_pts, 27700)
croy_ff_bng <- st_transform(croydon_ff_pts, 27700)
croydon_bng <- st_transform(croydon, 27700)
croy_ff_ppp <- as.ppp(croy_ff_bng)
croydon_owin <- as.owin(croydon_bng)
Window(croy_ff_ppp) <- croydon_owin

# Quadrat count analysis
Q <- quadratcount(croy_ff_ppp, nx= 6, ny=6)
plot(croy_ff_ppp, pch=20, cols="grey70", main=NULL); plot(Q, add=T)

```

## Spatial Density Analysis
```{r}
plot(rhohat(croy_ff_ppp, geo_hex_im))
```

## Kolmogorov-Smirnov Test for Spatial Distribution
```{r}

ks_result <- cdf.test(croy_ff_ppp, "x")
print(ks_result)

```

# Practical session - now it's your turn. Select a different type of point from Open Street Map (https://wiki.openstreetmap.org/wiki/Key:amenity) which may have an impact on children's health and investigate whether they cluster disproportionately around schools.


